Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-12-08 paper/manuscript.tex
draft of diversifying selection section

\begin{table}
\begin{fullwidth}
{\centering \include{figures/gamma_omega_modelcomparison}}
\caption{\label{tab:phydms}
Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature much better than a standard substitution model.
Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in Figure~\ref{fig:tree}.
Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
Both ExpCM models have a stringency parameter $>$1.
All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected, as determined using the approach in \citet{bloom2017identification}.
Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results of analyses of the same tree with additional substitution models.
}
\tabledata{\label{tabledata:phydms}Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites to eliminate the site specificity are in \texttt{modelcomparison.md}.}
\end{fullwidth}
\end{table}

@@

\subsection{Comparing selection in the lab to natural selection}
All the foregoing experiments measure the effects of mutations on viral replication \emph{in a T-cell line in the lab}.
But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
For instance, antibody pressure might increase the rate of evolution at some sites~\citep{wei2003antibody}, whereas pressure to mask certain epitopes~\citep{kwong2002hiv} might add constraint at other sites. 
Comparing selection in our experiments to natural selection can identify sites that are under such additional pressures during HIV's actual evolution in humans.
In contrast, standard methods for analyzing protein evolution that do not use experiments only test if sites are changing faster or slower than expected under neutral selection~\citep[so-called dN/dS tests;][]{yang2000codon,kosakovsky2005not,murrell2013fubar} -- something that is of less utility since we already know that Env is under strong functional constraint.

\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
\caption{\label{fig:divsel}
Sites in Env that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
We calculated the statistical evidence ($P$ value) that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
We then converted the $P$ values to $Q$ values to interpret them in the context of the false discovery rate~\citep{storey2003statistical}
{\bf (A)}
One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to red based on the strength of evidence that sites are evolving slower (blue) or faster (red) than expected given the BG505 experiments.
Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
{\bf (B)}
Sites of slower or faster than expected evolution given the BF520 experiments.
For both sets of experiments, sites that evolve faster or slower than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
{\bf (C)}
The results are similar regardless of whether the BG505 or BF520 experiments are used.
Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs.
All sites that evolve slower than expected for both experimental datasets are in Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}.
{\bf (D)} 
A large cluster of sites that evolve slower than expected is likely involved in Env's transition between open and closed conformational states.
Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form the hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
}
\figsupp[\label{figsupp:divselRSA}Relative solvent accessibilities of sites evolving faster or slower than expected.]{
Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} homologs, or whether they do fall into neither of these categories.
Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
\figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both homologs.
Specifically, it shows all sites with $\omega_r < 1$ at $Q < 0.05$ for the ExpCMs for \emph{both} BG505 and BF520.
For each site, the plots show the preferences averaged across replicates and re-scaled for each homolog, as well as the frequencies of amino acids in the clade A Env alignment. 
The $Q$-value indicated is the maximum of that for the two homologs.
}{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
\figdata{The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
\end{figure}

We determined whether each site in Env was evolving faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520.

>>>>

\begin{table}
\begin{fullwidth}
{\centering \include{figures/gamma_omega_modelcomparison}}
\caption{\label{tab:phydms}
Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature much better than a standard substitution model.
Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in Figure~\ref{fig:tree}.
Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
Both ExpCM models have a stringency parameter $>$1.
All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected at a false discovery rate of 0.05, as determined using the approach in \citet{bloom2017identification}.
Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results for additional substitution models.
}
\tabledata{\label{tabledata:phydms}Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites to eliminate the site specificity are in \texttt{modelcomparison.md}.}
\end{fullwidth}
\end{table}

@@

\subsection{Comparing selection in the lab to natural selection}
All the foregoing experiments measure the effects of mutations on viral replication in a T-cell line in the lab.
But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
For instance, antibody pressure might increase the rate of evolution at some sites~\citep{albert1990rapid,wei2003antibody,richman2003rapid}, whereas pressure to mask certain epitopes~\citep{kwong2002hiv} might add constraint at other sites. 
Comparing selection in our experiments to natural selection can identify sites that are under such additional pressures during HIV's actual evolution in humans.
%In contrast, standard methods for analyzing protein evolution that do not use experiments only test if sites are changing faster or slower than expected under neutral selection~\citep[so-called dN/dS tests;][]{yang2000codon,kosakovsky2005not,murrell2013fubar}.
% -- something that is of less utility since we already know that Env is under strong functional constraint.

\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/omegabysite_structural_analysis/omegabysite_structural_analysis.pdf}
\caption{\label{fig:divsel}
Sites in Env that evolve faster or slower in nature than expected given the functional constraints measured in the lab.
We calculated the statistical evidence ($P$ value) that each site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experimentally measured amino-acid preferences using the method of \citet{bloom2017identification}.
We then converted the $P$ values to $Q$ values to interpret them in the context of the false discovery rate~\citep{storey2003statistical}
{\bf (A)}
One monomer of the Env trimer~\citep[PDB 5FYL;][]{stewart2016trimeric} is colored from blue to red based on the strength of evidence that sites are evolving slower (blue) or faster (red) than expected given the BG505 experiments.
Sites where the rate of evolution is significantly different than expected at a false discovery rate of 0.05 are shown in spheres.
{\bf (B)}
Sites of slower or faster than expected evolution given the BF520 experiments.
For both sets of experiments, sites that evolve faster or slower than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).
{\bf (C)}
The results are similar regardless of whether the BG505 or BF520 experiments are used.
Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs.
All sites that evolve slower than expected for both experimental datasets are in Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}.
{\bf (D)} 
A large cluster of sites that evolve slower than expected is likely involved in Env's transition between open and closed conformational states.
Gray boxes indicate sites that \citet[PDB 5VN3]{ozorowski2017open} proposed form the hydrophobic network that regulates the conformational change; blue boxes and sticks indicate sites that evolve slower than expected.
All analyses used the phylogenetic tree in Figure~\ref{fig:tree}.
}
\figsupp[\label{figsupp:divselRSA}Relative solvent accessibilities of sites evolving faster or slower than expected.]{
Sites are grouped by whether they have $\omega_r > 1$ (diversifying selection) or $\omega_r < 1$ (purifying selection) at $Q < 0.05$ in \emph{both} homologs, or whether they do fall into neither of these categories.
Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.
As can be seen from these box plots with overlaid points for each site, sites of both diversifying and purifying selection tend to have higher relative solvent accessibility than other sites.
}{\includegraphics[width=0.5\textwidth]{figures/selsites_rsa.pdf}}
\figsupp[\label{figsupp:slower}Amino-acid preferences and alignment frequencies of sites that evolve slower than expected.]{
This plot shows all sites that are evolving more slowly than expected in natural sequences given the preferences measured in both homologs.
Specifically, it shows all sites with $\omega_r < 1$ at $Q < 0.05$ for the ExpCMs for \emph{both} BG505 and BF520.
For each site, the plots show the preferences averaged across replicates and re-scaled for each homolog, as well as the frequencies of amino acids in the clade A Env alignment. 
The $Q$-value indicated is the maximum of that for the two homologs.
}{\includegraphics[width=\textwidth]{figures/purifying_selsites.pdf}}
\figdata{The $\omega_r$ and $Q$-values are in \texttt{merged\_omegabysite.csv}.}
\end{figure}

We determined whether each site in Env evolves faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520. 
The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
As shown in Table~\ref{tab:phydms}, the standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution.
This finding is only of modest interest, since it is well known that Env is under functional constraint and so has many sites that evolve non-neutrally.
In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).

The sites that evolve slower or faster than expected from the experiments are shown in Figure~\ref{fig:divsel}A,B.
The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (Figure~\ref{fig:divsel}C).
The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, meaning that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in Figure~\ref{fig:tree}.

What causes some sites to evolve faster or slower in nature than expected from the experiments?
The answer in both cases is likely to be immune selection.
Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}).
It is known that Env's escape from autologous neutralizing antibodies often involves amino-acid substitutions in surface-exposed regions~\citep{moore2009specificity}, including at many of the sites that evolve faster than expected.
Since our deep mutational scanning did not impose antibody pressure, sites where substitutions are antibody-driven will evolve faster in nature than expected from the experiments.
Interestingly, immune selection also offers a plausible explanation for the sites that evolve \emph{slower} than expected.
In addition to escaping immunity via substitutions at antibody-binding footprints, Env is notorious for employing a range of more general strategies to reduce its susceptibility to antibodies.
These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
Sites that contribute to such immune-evasion strategies will be under a constraint in nature that is not present in our experiments, since the experiments do not select for immune evasion.
%Indeed, many of the sites that evolve slower than expected appear to be involved in these general immune-evasion strategies.
For instance, sites of N-linked glycans tend to be more conserved in nature than the experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C), as are sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.
